---
title: "CYVD Model -- Sensitivity Analysis"
author: "SYM"
date: "2/22/2022"
output:
  html_document:
    self_contained: no
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives/Introduction
This model tracks a population of squash bugs that vary in their symbiont and pathogen infection status in a field over the course of a single squash growing season. We specifically look at how varying traits impacted by symbiont status affects pathogen transmission. 
Objectives: 
Identify whether vector traits influenced by symbiont status are important in determining plant infection dynamics.

Identify the role of aposymbiotic bugs in propagating transmission and determine whether symbiont prevalence has an effect. 

This script runs the updated CYVD model and global sensitivity analysis on 3 parameters of interest:
apo bug clearance rate, apo bug transmission rate, and probability of symbiont acquisition (symbiont prevalence).

## Model and justification
A model diagram with equations can be found in our paper in Ecological Modelling (doi XXXXXXX). All equations and state variables are also available in the C file. 

## Data source(s)
Our model makes use of data that are already published and publicly available. All relevant citations can be found in Table 1 of our paper.

Data sets utilized for this research are as follows: 
Data on insect maturation rates: https://doi.org/10.3389/fmicb.2021.719112 
Acevedo, T. S., G. P. Fricker, J. R. Garcia, T. Alcaide, A. Berasategui, K. S. Stoy, and N. M. Gerardo. 2021. The Importance of Environmentally Acquired Bacterial Symbionts for the Squash Bug (Anasa tristis), a Significant Agricultural Pest. Frontiers in Microbiology 12:719112.
Data on overwintering adults: https://shareok.org/handle/11244/300885?show=full 
Eiben, J. A. 2004. Characterizing the Overwintering and Emergence Behaviors of the Adult Squash Bug, Anasa tristis (Degeer). Albright College.
Data on insect survival:  https://www.proquest.com/docview/303565822/abstract/B46754349D354E12PQ/1 
Fielding, D. J. 1988. Phenology and population dynamics of the squash bug, Anasa tristis . Ph.D., University of Illinois at Urbana-Champaign, United States -- Illinois.
Data on impact of symbiont status on insect vector traits: https://doi.org/10.1128/AEM.01550-21  
Mendiola, S. Y., K. S. Stoy, S. DiSalvo, C. L. Wynn, D. J. Civitello, and N. M. Gerardo. 2022. Competitive Exclusion of Phytopathogenic Serratia marcescens from Squash Bug Vectors by the Gut Endosymbiont  Caballeronia . Applied and Environmental Microbiology 88:e01550-21.
Data on adult insect reproduction: https://doi.org/10.1007/s10682-021-10120-1
Villa, S. M., D. D. Han, M. E. Jordan, and N. M. Gerardo. 2021. The resilience of reproductive interference. Evolutionary Ecology 35:537–553.
 
## Scripts
```{r load required packages}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot()) # Makes the ggplot look nicer (says Dave)
library(tidyverse)
library(tidyr)
library(patchwork)
library(dplyr)
library(deSolve)
library(caTools) #this package is for AUC calculations
library(sensobol)
library(data.table)
library(foreach)
library(parallel)
library(doParallel)

```

##Load model and compile from C
```{r load_model}
# compile model from C definition
try(dyn.unload("C:/Users/sandr/Dropbox/squash_bug_model/Squash_bug_apographs3.dll")) # unLoad dll
system("R CMD SHLIB C:/Users/sandr/Dropbox/squash_bug_model/Squash_bug_apographs3.c")
dyn.load("C:/Users/sandr/Dropbox/squash_bug_model/Squash_bug_apographs3.dll") # load dll

```
## Model setup
```{r set up model}
#### low transmission ####
Inits = c(OA = 9, E = 0, L1 = 0, L2 = 0, L3 = 0, L4 = 0, L5 = 0, A = 0, L2o = 0, L3o = 0, L4o = 0, L5o = 0, Ao = 0, L2I = 0, L3I = 0, L4I = 0, L5I = 0, AI = 0, L2oI = 0, L3oI = 0, L4oI = 0, L5oI = 0, AoI = 0, P_I = 0, Apo_PI = 0, Sym_PI = 0, AO_PI = 0, OAI = 1)
pars_low = c(d_A = 1/60, l = 0.05, b = 10, K_E = 250, 
         d_E = 1/7, m_E = 1/7, # Development rates are rough estimates from Acevedo et al
         d_1 = 1/4, m_1 = 1/4, # Mortality rates just correspond to 50% survival through the stages
         d_2 = 1/4, m_2 = 1/4,
         d_3 = 1/4, m_3 = 1/4,
         d_4 = 1/4, m_4 = 1/4,
         d_5 = 1/16, m_5 = 1/16,
         
         p = 0.9, #set this anywhere between .9 to 1, we will vary the range later
         d_2o = 1/4, m_2o = 1/4,
         d_3o = 1/7, m_3o = 1/7,
         d_4o = 1/13, m_4o = 1/13,
         d_5o = 1/17, m_5o = 1/17,
         d_Ao = 1/60,
         
         c = 1/3, B_pb = 0.025,
         c_o = 1/10, B_bp = 0.01,
         B_o = 10, P0 = 1
         )
timespan = 180:270 #field season by Julian date

sim1 = lsoda(y = Inits, times=timespan, parms = pars_low, func="derivs", dllname = "Squash_bug_apographs3", initfunc="initmod")
```

```{r plot population dynamics}
#create a data frame from simulation output for graphing 

preds = dim(sim1)[1]

Stage = rep(colnames(sim1)[-1], each=preds)
Jul.date = rep(sim1[,1], times = length(colnames(sim1)[-1]))
Counts = c(sim1[,-1])

prediction.df = data.frame(Stage, Jul.date, Counts)


p.plot <- ggplot(data=prediction.df, aes(x=Jul.date, y=Counts, group=Stage, colour=Stage)) + 
         geom_line(size = 1.4) +
         ylim(c(0, 200)) + xlim(c(175, 275)) +
         scale_fill_manual(name = "Life Stage", labels = c('eggs', 'first', 'second', 'third', 'fourth', 'fifth', 'adults'))+
        labs(y = "Bugs per plant", x = "Julian Date") 
p.plot

p.preds <- ggplot(data=prediction.df, aes(x=Jul.date, y=Counts, group=Stage, colour=Stage)) + geom_line(size = 1.4)+
          ylim(c(0,90)) + xlim(c(175, 275)) 
          
p.preds

```

### sensobol aesthetic mods
```{r set new plot functions for sensobol}
#function sandra_plot_scatter alters the graph output of the sensobol function to be more visually pleasing to Sandra
#run this before running sensobol 

library(ggplot2)

plot_names <- c('B_o' = "\u03BB",
                'p' = "p",
                'c_o' = "\u03B3^-1")

axis_titles <- data.frame(
  variable = c("B_o", "c_o", "p"),
  axis_title = c("Fold increase in \n apo transmission", "Fold increase in \n apo clearance time", "Probability of \n symbiont acquisition")
)


sandra_plot_scatter = function (data, N, Y, params, method = "point", size = 0.7, 
  alpha = 0.2) 
{
  value <- y <- NULL
  minY <- min(Y)
  print(minY)
  dt <- data.table::data.table(cbind(data, Y))[1:N]
  colnames(dt)[length(colnames(dt))] <- "y"
  out <- data.table::melt(dt, measure.vars = params)
  gg <- ggplot2::ggplot(out, ggplot2::aes(value, y)) + 
    ggplot2::facet_wrap(~variable, scales = "free_x", labeller = as_labeller(plot_names, default = label_parsed))  + 
    ggplot2::labs(x = NULL, y = "Area under infection prevalence curve") + 
    ggplot2::theme_bw() + 
    ggplot2::theme(panel.grid.major = ggplot2::element_blank(),
                   panel.spacing = unit(1.25, "lines"), #add more space between panels
                  panel.grid.minor = ggplot2::element_blank(), legend.background = ggplot2::element_rect(fill = "transparent", 
                  color = NA), legend.key = ggplot2::element_rect(fill = "transparent", 
                  color = NA), strip.background = ggplot2::element_rect(fill = "white"), 
                  strip.text = element_text(size = 14), legend.position = "top",
                  plot.margin = unit(c(.5,.1,.75,.5), "cm"),
                  axis.title=element_text(size=14), axis.text = element_text(size = 12)) + 
    ggplot2::geom_text(data = axis_titles, aes(label = axis_title), size = 4.5, 
                       x= c(min(dt$B_o) + diff(range(dt$B_o))/2, min(dt$c_o) + diff(range(dt$c_o))/2, min(dt$p) + diff(range(dt$p))/2) ,  #centers x axis labels for individual graphs, variables hardcoded, need to change if using diff ones
                       y= minY - diff(range(Y))/4, color='black') + # need to change y to the minimal value - some divider x Y's range
    coord_cartesian(clip="off")
  if (method == "point") {
    gg <- gg + ggplot2::geom_point(size = size, alpha = alpha) + 
      ggplot2::stat_summary_bin(fun = "mean", geom = "point", 
        colour = "red", size = 1.2)
  }
  else if (method == "bin") {
    gg <- gg +  ggplot2::geom_hex(bins = 35) + scale_fill_viridis_c() + #scale_colour_gradient2(palette = "viridis") +
      ggplot2::stat_summary_bin(fun = "mean", geom = "line", colour = "red", size = 1.2)
  }
  else {
    stop("Method should be either point or bin")
  }
  return(gg)
}


#another mod to graphing defaults 

library(RColorBrewer)

sandra_plot_multiscatter = function (data, N, Y, params, smpl = NULL) 
{
    xvar <- yvar <- x <- y <- NULL
    dt <- data.table::data.table(data)
    out <- t(utils::combn(params, 2))
    da <- list()
    for (i in 1:nrow(out)) {
        cols <- out[i, ]
        da[[i]] <- cbind(dt[1:N, .SD, .SDcols = (cols)], cols[1], 
            cols[2], Y[1:N])
        data.table::setnames(da[[i]], colnames(da[[i]]), c("xvar", 
            "yvar", "x", "y", "output"))
    }
    output <- data.table::rbindlist(da)
    if (is.null(smpl) == FALSE) {
        if (is.numeric(smpl) == FALSE) {
            stop("smpl should be a number")
        }
        else {
            output <- output[, .SD[sample(.N, min(smpl, .N))], 
                by = list(x, y)]
        }
    }
    gg <- ggplot2::ggplot(output, ggplot2::aes(xvar, yvar, color = output)) + 
        ggplot2::geom_point(size = 0.5) + ggplot2::scale_colour_gradientn(colours = grDevices::hcl.colors(7, palette = "viridis")) + 
        ggplot2::scale_x_continuous(breaks = scales::pretty_breaks(n = 3)) + 
        ggplot2::scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) + 
        ggplot2::facet_wrap(x ~ y, scales = "free") + ggplot2::theme_bw() + 
        ggplot2::labs(x = "", y = "") + ggplot2::theme(panel.grid.major = ggplot2::element_blank(), 
        panel.grid.minor = ggplot2::element_blank(), legend.background = ggplot2::element_rect(fill = "transparent", 
            color = NA), legend.key = ggplot2::element_rect(fill = "transparent", 
            color = NA), strip.background = ggplot2::element_rect(fill = "white"), 
        legend.position = "top")
    return(gg)
}

```

```{r sensobol graphs low transmission setting}
N <- 2^13
params <- c( "B_o", "p", "c_o") # Choose your parameters of interest
matrices <- c("A", "B", "AB", "BA")
first <- total <- "azzini"
order <- "second"
R<-10^3 
type<-"percent"
conf<-0.95

mat.low <- sobol_matrices(matrices=matrices, N = N, params = params, order = order)

# Choose their ranges
mat.low[, "B_o"] <- qunif(mat.low[, "B_o"], 10, 1000)
mat.low[, "p"] <- qunif(mat.low[, "p"], .9, 1)
mat.low[, "c_o"] <- qunif(mat.low[, "c_o"], .1, 1)


sobol_fxn = function(B_o,p,c_o) { # Runs the model and calculates area under the prevalence curve
  pars_low["B_o"] = B_o; pars_low["p"] = p; pars_low["c_o"] = c_o;
  sim.low = lsoda(y = Inits, times=timespan, parms = pars_low, func="derivs", dllname = "Squash_bug_apographs3", initfunc="initmod", maxsteps = 500000)
  return(list(c(auc_PI = trapz(sim.low[,"time"], sim.low[,"P_I"]), auc_ApoPI = trapz(sim.low[,"time"], sim.low[,"Apo_PI"]), auc_SymPI = trapz(sim.low[,"time"], sim.low[,"Sym_PI"])))) 
} 

sobol_fxn(B_o = pars_low["B_o"],p = pars_low["p"], c_o = pars_low["c_o"])

sobol_mapply_fxn = function(dt){
  return(mapply(sobol_fxn, dt[, 1], dt[, 2], dt[, 3])) }

# mat = our inputs
# result = our model outputs
result_low = sobol_mapply_fxn(mat.low)
df.low = as.data.frame(do.call(rbind, result_low))
full.dt.low = data.table(mat.low, df.low) 

#for inverting the clearance rate 
mat.low.fin <- mat.low %>%
  as.data.frame(mat.low) %>%
  mutate(c_o = (1/c_o))

full.dt.low.fin = data.table(mat.low.fin, df.low)

plot_uncertainty(Y = full.dt.low.fin$auc_PI, N = N) + labs(y = "Mean prevalence", x = "y")

#sensitivity analysis graphs

low.mat <- sandra_plot_scatter(data = mat.low.fin, N = N, Y = full.dt.low.fin$auc_PI, params = params, method = "bin") 

#plots all plant infections
low.mat

# In labels, top = X, bottom = Y
sandra_plot_multiscatter(data = mat.low.fin, N = N, Y = df.low$auc_PI, params = params, smpl = 2^13)

ind <- sobol_indices(matrices = matrices, Y = df.low$auc_PI, N = N, params = params, 
                     first = first, total = total, order = order, boot = TRUE, R = R,
                     parallel = "no", type = type, conf = conf)

cols <- colnames(ind$results)[1:5]
ind$results[, (cols):= round(.SD, 3), .SDcols = (cols)]
ind

toplot.low <- subset(ind$results, parameters=="B_o" | parameters=="c_o" | parameters=="p")
toplot.low

```

```{r sensobol graphs high transmission}
#### high transmission parameters####
Inits = c(OA = 9, E = 0, L1 = 0, L2 = 0, L3 = 0, L4 = 0, L5 = 0, A = 0, L2o = 0, L3o = 0, L4o = 0, L5o = 0, Ao = 0, L2I = 0, L3I = 0, L4I = 0, L5I = 0, AI = 0, L2oI = 0, L3oI = 0, L4oI = 0, L5oI = 0, AoI = 0, P_I = 0, Apo_PI = 0, Sym_PI = 0, OA_PI = 0, OAI = 1)

pars_hi = c(d_A = 1/60, l = 0.05, b = 10, K_E = 250, 
         d_E = 1/7, m_E = 1/7, # Development rates are rough estimates from Tarik's dataset
         d_1 = 1/4, m_1 = 1/4, # Mortality rates just correspond to 50% survival through the stages
         d_2 = 1/4, m_2 = 1/4,
         d_3 = 1/4, m_3 = 1/4,
         d_4 = 1/4, m_4 = 1/4,
         d_5 = 1/16, m_5 = 1/16,
         
         p = 0.9,
         d_2o = 1/4, m_2o = 1/4,
         d_3o = 1/7, m_3o = 1/7,
         d_4o = 1/13, m_4o = 1/13,
         d_5o = 1/17, m_5o = 1/17,
         d_Ao = 1/60,
         
         c = 1/3, B_pb = 0.025,
         c_o = 1/10, B_bp = 0.1, #note all params are the same as before except for B_bp
         B_o = 20, P0 = 1
         )
timespan = 180:270

sim.hi = lsoda(y = Inits, times=timespan, parms = pars_hi, func="derivs", dllname = "Squash_bug_apographs3", initfunc="initmod")


N <- 2^13
params <- c( "B_o", "p", "c_o") # Choose your parameters of interest
matrices <- c("A", "B", "AB", "BA")
first <- total <- "azzini"
order <- "second"
R<-10^3 
type<-"percent"
conf<-0.95

mat.hi <- sobol_matrices(matrices=matrices, N = N, params = params, order = order)

# Choose their ranges
mat.hi[, "B_o"] <- qunif(mat.hi[, "B_o"], 10, 1000)
mat.hi[, "p"] <- qunif(mat.hi[, "p"], .9, 1)
mat.hi[, "c_o"] <- qunif(mat.hi[, "c_o"], 1/10, 1)


sobol_fxn_hi = function(B_o,p,c_o) { # Runs the model and calculates area under the prevalence curve
  pars_hi["B_o"] = B_o; pars_hi["p"] = p; pars_hi["c_o"] = c_o;
  sim.hi = lsoda(y = Inits, times=timespan, parms = pars_hi, func="derivs", dllname = "Squash_bug_apographs3", initfunc="initmod", maxsteps = 500000)
  return(list(c(auc_PI = trapz(sim.hi[,"time"], sim.hi[,"P_I"]), auc_ApoPI = trapz(sim.hi[,"time"], sim.hi[,"Apo_PI"]), auc_SymPI = trapz(sim.hi[,"time"], sim.hi[,"Sym_PI"])))) 
} 

sobol_fxn_hi(B_o = pars_hi["B_o"],p = pars_hi["p"], c_o = pars_hi["c_o"])

sobol_mapply_fxn_hi = function(dt){
  return(mapply(sobol_fxn_hi, dt[, 1], dt[, 2], dt[, 3])) 
  
}

# mat = our inputs
# result = our model outputs
result_hi = sobol_mapply_fxn_hi(mat.hi)
df.hi = as.data.frame(do.call(rbind, result_hi))
full.dt.hi = data.table(mat.hi, df.hi) 

####for inverting the clearance rate  ####
mat.hi.fin <- mat.hi %>%
  as.data.frame(mat.hi) %>%
  mutate(c_o = (1/c_o))

full.dt.hi.fin = data.table(mat.hi.fin, df.hi)

#### plots ####
plot_uncertainty(Y = full.dt.hi.fin$auc_PI, N = N) + labs(y = "Mean prevalence", x = "y")

hi.mat <- sandra_plot_scatter(data = mat.hi.fin, N = N, Y = full.dt.hi.fin$auc_PI, params = params, method = "bin") #plots all plant infections
hi.mat

# In labels, top = X, bottom = Y
sandra_plot_multiscatter(data = mat.hi.fin, N = N, Y = df.hi$auc_PI, params = params, smpl = 2^13)

ind.hi <- sobol_indices(matrices = matrices, Y = df.hi$auc_PI, N = N, params = params, 
                     first = first, total = total, order = order, boot = TRUE, R = R,
                     parallel = "no", type = type, conf = conf)

cols <- colnames(ind.hi$results)[1:5]
ind.hi$results[, (cols):= round(.SD, 3), .SDcols = (cols)]
ind.hi

toplot.hi <- subset(ind.hi$results, parameters=="B_o" | parameters=="c_o" | parameters=="p")
toplot.hi

```

```{r sobol indices graphs}
#look at plots
plot(ind.hi)
plot(ind)

#make plots nicer
ind.low <- ggplot(toplot.low, aes(x=parameters, y=original, fill=sensitivity)) + 
  geom_bar(stat="identity", color="black", position=position_dodge()) +
  scale_fill_manual(name = "Effects", labels = c("Individual", "Total"), values=c("grey",
                             "gray42")) +
  geom_linerange(aes(ymin=original-std.error, ymax=original+std.error), linewidth= 1.5,
                 position=position_dodge(.9)) +
   theme(legend.position = "top", legend.justification = "center") +
  scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_x_discrete(labels=c("Apo Transmission \n Rate", "Apo Clearance \n Rate", "Probability of \n Symbiont Acquisition")) +
  labs(y = "Index Value", x = "")

ind.low

ind.hi <- ggplot(toplot.hi, aes(x=parameters, y=original, fill=sensitivity)) + 
  geom_bar(stat="identity", color="black", position=position_dodge()) +
  scale_fill_manual(name = "Effects", labels = c("Individual", "Total"), values=c("grey",
                             "gray42")) +
  geom_linerange(aes(ymin=original-std.error, ymax=original+std.error), linewidth= 1.5,
                 position=position_dodge(.9)) +
  theme(legend.position = "top", legend.justification = "center") +
  scale_y_continuous(limits = c(0, 1), expand = c(0,0)) +
  scale_x_discrete(labels=c("Apo Transmission \n Rate", "Apo Clearance \n Rate", "Probability of \n Symbiont Acquisition")) +
  labs(y = "Index Value", x = "")

ind.hi
```

##putting all graphs together
```{r fig.height = 11, fig.width=16}

fig <- plot_grid(NULL, low.mat, NULL, ind.low, NULL, NULL, NULL, NULL, NULL, hi.mat, NULL, ind.hi, 
                 labels = c('', 'A', '', 'B', '', '', '', '', '', 'C', '', 'D'), label_size = 20, axis = "tlbr",
  align = 'vh', nrow = 4, ncol = 4, rel_widths = c(.09, 1.35, .1, 1), rel_heights= c(1, .05, 1, .05), 
  scale = c(1, 1, 1, 1.1, 1, 1, 1, 1.1)) 
fig + annotate("text", x = .0125, y = .75, angle = "90", size = 9.5, fontface = "bold", label = "Low Transmission") + annotate("text", x = .0125, y = .25, angle = "90", size = 9.5, fontface = "bold", label = "High Transmission")

#save figure 
tiff(file="sensobol_fig3.tiff",width=18, height=12, units="in", res=300)
fig + annotate("text", x = .0125, y = .75, angle = "90", size = 9.5, fontface = "bold", label = "Low Transmission") + annotate("text", x = .0125, y = .25, angle = "90", size = 9.5, fontface = "bold", label = "High Transmission")
dev.off()

```

